Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-04 paper.tex
Tightening language more.

\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
  Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
  NVIDIA, by contrast uses fairness in line with our language. From Tesla architecture: \newline
  "A scoreboard qualifies each warp for issue each cycle. The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue. Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
  From DecoupledLookback: \newline
  "Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
  "Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
}

@@

\subsection{Ideal Scan Architecture}
\emph{Chained-Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy. By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the hardware scheduler to dynamically manage occupancy. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.

@@

Given these constraints, oversubscription and \emph{Chained-Scan} emerge as the more portable and adaptable architecture. By leveraging the hardware scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and certain ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline ``Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated'' \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.} \john{I don't have an opinion on this, you're familiar with this and I'm not, but this seems like a good footnote. Or we can ask Tyler.}

@@

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

@@

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. \john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is? Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
  allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---Not Ready $\rightarrow$ Ready $\rightarrow$ Inclusive---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Fallback:}
        \begin{itemize}
          \item If the blocking tile is still \emph{Not Ready} after waiting, the workgroup initiates a fallback operation.
          \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements. This process involves a \emph{workgroup-wide reduction},\footnote{See footnote NUMBER.} utilizing additional shared memory to aggregate partial subgroup results.
          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
          \item As this update attempt is performed atomically, the workgroup simultaneously reads back the blocking tile's most recent state:
                \begin{itemize}
                  \item If that most recent state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
                  \item Otherwise, the first thread ingests the fallback value.
                        \begin{itemize}
                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the lookback. The first thread proceeds to step 4.
                            \item Otherwise, the first thread updates \emph{current tile} to the preceding tile, returning to the lookback process in step~2.
                        \end{itemize}
                \end{itemize}
        \end{itemize}

  \item \textbf{Finalization:}
        \begin{itemize}
          \item To prioritize inter-workgroup reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to apply it in their \emph{workgroup-wide} and local scan computations.
          \item Finally, the first thread sets the control flag to \emph{Unlocked}, signaling the rest of the workgroup that \emph{Chained Propagation} has completed, and allowing the final write phase to begin.
        \end{itemize}
\end{enumerate}

\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance. For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance. Perhaps worth pointing out. Does the flag-unset have any ordering constraints in terms of correctness?}
\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}

>>>>

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce then Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy. By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the workgroup scheduler to dynamically manage occupancy. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. By leveraging the workgroup scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.

@@

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery}\footnote{\emph{Timeout detection and recovery} is} (TDR) and crashing the host program.

@@

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. \john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is? Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{} subgroup scans,
  allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone~\cite{} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Fallback:}
        \begin{itemize}
          \item If the blocking tile is still \emph{Not Ready} after waiting, the workgroup initiates a fallback operation.
          \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements. This process involves a \emph{workgroup-wide reduction},\footnote{See footnote NUMBER.} utilizing additional shared memory to aggregate partial subgroup results.
          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
          \item As this update attempt is performed atomically, the workgroup simultaneously reads back the blocking tile's most recent state:
                \begin{itemize}
                  \item If that most recent state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
                  \item Otherwise, the first thread ingests the fallback value.
                        \begin{itemize}
                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the \emph{Chained Propagation}. The first thread proceeds to step 4.
                            \item Otherwise, the first thread updates \emph{current tile} to the preceding tile, returning to the lookback process in step~2.
                        \end{itemize}
                \end{itemize}
        \end{itemize}

  \item \textbf{Finalization:}
        \begin{itemize}
          \item To prioritize inter-workgroup reduction propagation, the first thread immediately adds the local reduction to \emph{previous reduction}, posts the inclusive reduction to global memory, and transitions the tile state to \emph{Inclusive}.
          \item It then writes \emph{previous reduction} to shared memory, allowing all threads to combine it with their \emph{workgroup-wide} and private scan computations.
          \item Finally, the first thread sets the control flag to \emph{Unlocked}, signaling the rest of the workgroup that \emph{Chained Propagation} has completed, and allowing the final write phase to begin.
        \end{itemize}
\end{enumerate}

\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance. For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance. Perhaps worth pointing out. Does the flag-unset have any ordering constraints in terms of correctness?}
\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
\thomas{No furher revisions required/OK to remove comments?}.
